Connecting R with D3 for dynamic graphics, to explore multivariate data with tours

### 2018-02-19 paper-tourrd3.Rmd
Small changes/fixes in the text.

---
title: Connecting R with D3 for dynamic graphics, to explore multivariate data with tours
author:
  - name: Michael Kipp
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  mkipp271@gmail.com
  - name: Dianne Cook
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu 
  - name: Ursula Laa
    affiliation: Monash University
    address:
    - Department of Physics
    email:  ursula.laa@monash.edu
bibliography: references.bib
link-citations: true
fig_caption: yes
abstract: >
  The tourr package in R has several algorithms and displays for showing multivariate data as a sequence of low-dimensional projections. It can display as a movie but has no capacity for interaction, such as stop/go, change tour type, drop/add variables. The tourrGui  package provides these sorts of controls, but the interface is programmed with the dated RGtk2 package. This work explores using custom messages to pass data from R to D3  for viewing, using the shiny framework.  
preamble: >
  % Any extra latex you need in the preamble
output: rticles::rjournal_article
---

@@

These interactive controls were provided with the tourrGui package [@tourrGui], with was programmed with the RGtk2 package [@RGtk2]. This is not the toolkit of choice today, and has been superceded with primarily web-capable tools, like shiny [@shiny]. To display dynamic graphics though, is not straight-forward. This paper explains how to use D3 [@D3] as the display engine, in a shiny GUI, using custom message passing between server and client.

@@

The new implementation of the tour uses D3 to render each projection step returned by R, focusing on 2D projections as a test case. It does this by drawing and re-drawing a scatterplot with dots (or circles in D3 language) and providing SVG objects for the web browser to render.

There two functions provided by the Shiny framework to transport data be- tween R and javascript: `session$sendCustomMessage()` in R, and the corresponding `Shiny.addCustomMessageHandler()` in Javascript. Whenever the former is executed in R, the latter function will execute a code block in JS. There are many examples of such functions being used to pass arbitrary data from an R app to a JS front-end, few examples exist of this basic functionality to update a D3 animation in real-time.

The data format expected by D3 is in JSON format, which combines two basic program- ming paradigms: a collection of name/value pairs, and an ordered list of values. R's preferred data formats include data frames, vectors and matrices. Every time a new projection has been calculated with the tour path, the resulting matrix needs to be converted to JSON and sent to D3. The code to send the D3 data looks like this:

@@

This code is from the observe environment from the `server.R` file. It converts the projection matrix to JSON format, and sends it to JavaScript with the id data. When parsed in D3 by its `data()` method, it is converted back into a logical 2D array where the columns are queried first, then the rows. If column names are included in the JSON, the column indices are strings; otherwise they are integers starting from 1.
All of the code required to render the scatterplots and legends, along with colours, is JavaScript code in the file `d3anim.js`. In particular, the data from R is handled with the following code:

@@

The D3 canvas makes for smooth drawing and re-drawing of the data projections. Adding a graphical user interface (GUI) around the display is straightforward with the shiny package, e.g. control elements such as stop/go, increase/decrease speed, change tour type, add/remove variables from the mix.

The main disadvantage is that the speed is inconsistent, as server and client play tag to keep up with each other, and the display cannot handle many observations. Noticeable slow down was oberved with 2000 points.

@@

The `sendCustomMessage` function provides a way to share a tour path with the D3 renderer, and embed it in a shiny GUI providing controls such as top/go, increase/decrease speed, change tour type, add/remove variables. However, the approach can result in inconsistent speed which appears as jerkiness in the motion, and it isn't suitable for large numbers of observations.

>>>>

---
title: Connecting R with D3 for dynamic graphics, to explore multivariate data with tours
author:
  - name: Michael Kipp
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  mkipp271@gmail.com
  - name: Dianne Cook
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu 
  - name: Ursula Laa
    affiliation: Monash University
    address:
    - Department of Physics
    email:  ursula.laa@monash.edu
bibliography: references.bib
link-citations: true
fig_caption: yes
abstract: >
  The tourr package in R has several algorithms and displays for showing multivariate data as a sequence of low-dimensional projections. It can display as a movie but has no capacity for interaction, such as stop/go, change tour type, drop/add variables. The tourrGui  package provides these sorts of controls, but the interface is programmed with the dated RGtk2 package. This work explores using custom messages to pass data from R to D3  for viewing, using the Shiny framework.  
preamble: >
  % Any extra latex you need in the preamble
output: rticles::rjournal_article
---

@@

These interactive controls were provided with the tourrGui package [@tourrGui], with was programmed with the RGtk2 package [@RGtk2]. This is not the toolkit of choice today, and has been superceded with primarily web-capable tools, like Shiny [@shiny]. To display dynamic graphics though, is not straight-forward. This paper explains how to use D3 [@D3] as the display engine, in a Shiny GUI, using custom message passing between server and client.

@@

The new implementation of the tour interface uses D3 to render each projection step returned by R, focusing on 2D projections as a test case. It does this by drawing and re-drawing a scatterplot with dots (or circles in D3 language) and providing SVG objects for the web browser to render.

There are two functions provided by the Shiny framework to transport data between R and JavaScript: `session$sendCustomMessage()` in R, and the corresponding `Shiny.addCustomMessageHandler()` in JavaScript. Whenever the former is executed in R, the latter function will execute a code block in JS. There are many examples of such functions being used to pass arbitrary data from an R app to a JS front-end, few examples exist of this basic functionality to update a D3 animation in real-time.

The data format expected by D3 is in JSON format, which combines two basic programming paradigms: a collection of name/value pairs, and an ordered list of values. R's preferred data formats include data frames, vectors and matrices. Every time a new projection has been calculated with the tour path, the resulting matrix needs to be converted to JSON and sent to D3. The code to send the D3 data looks like this:

@@

This code is from the observe environment from the `server.R` file. It converts the matrix of projected data points to JSON format, and sends it to JavaScript with the id data. When parsed in D3 by its `data()` method, it is converted back into a logical 2D array where the columns are queried first, then the rows. If column names are included in the JSON, the column indices are strings; otherwise they are integers starting from 1.
All of the code required to render the scatterplots and legends, along with colours, is JavaScript code in the file `d3anim.js`. In particular, the data from R is handled with the following code:

@@

The D3 canvas makes for smooth drawing and re-drawing of the data projections. Adding a graphical user interface (GUI) around the display is straightforward with the Shiny package, e.g. control elements such as stop/go, increase/decrease speed, change tour type, add/remove variables from the mix.

The main disadvantage is that the speed is inconsistent, as server and client play tag to keep up with each other, and the display cannot handle many observations. Noticeable slow down was oberved with 2000 points, the main reason being the rendering time required for the large number of SVG circle elements. The situation can be improved when using a single HTML5 canvas element to draw the scatter points, significantly reducing the rendering time.

@@

The `sendCustomMessage` function provides a way to share a tour path with the D3 renderer, and embed it in a Shiny GUI providing controls such as stop/go, increase/decrease speed, change tour type, add/remove variables. However, the approach can result in inconsistent speed which appears as jerkiness in the motion, and it isn't suitable for large numbers of observations.

### 2018-02-19 paper-tourrd3.tex
Small changes/fixes in the text.

\abstract{%
The tourr package in R has several algorithms and displays for showing
multivariate data as a sequence of low-dimensional projections. It can
display as a movie but has no capacity for interaction, such as stop/go,
change tour type, drop/add variables. The tourrGui package provides
these sorts of controls, but the interface is programmed with the dated
RGtk2 package. This work explores using custom messages to pass data
from R to D3 for viewing, using the shiny framework.
}

@@

The tour algorithm (Cook et al. \protect\hyperlink{ref-gt_pp}{1995};
Cook et al. \protect\hyperlink{ref-gt_pp_mc}{2007}) is a way of
systematically generating and displaying projections of high-dimensional
spaces in order for the viewer to examine the multivariate distribution
of data. It can do this either randomly, or by picking projections
judged interesting according to some criterion or index function. The
tourr package (Wickham et al. \protect\hyperlink{ref-tourr}{2011})
provides the computing and display in R (R Core Team
\protect\hyperlink{ref-R}{2018}; Ihaka and Gentleman
\protect\hyperlink{ref-ihaka:1996}{1996}) to make several types of
tours, grand, guided, little and local. The projection dimension can be
chosen between one and the number of variables in the data. The display,
though, has no capacity for interaction. The viewer can watch the tour
like a movie, but not pause it and restart, or change tour type, or
number of variables.

These interactive controls were provided with the tourrGui package
(Huang, Cook, and Wickham \protect\hyperlink{ref-tourrGui}{2012}), with
was programmed with the RGtk2 package (Lawrence and Temple Lang
\protect\hyperlink{ref-RGtk2}{2010}). This is not the toolkit of choice
today, and has been superceded with primarily web-capable tools, like
shiny (Chang et al. \protect\hyperlink{ref-shiny}{2017}). To display
dynamic graphics though, is not straight-forward. This paper explains
how to use D3 (Bostock, Ogievetsky, and Heer
\protect\hyperlink{ref-D3}{2011}) as the display engine, in a shiny GUI,
using custom message passing between server and client.

@@

The \pkg{tourr} package (Wickham et al.
\protect\hyperlink{ref-tourr}{2011}) is an R implementation of the tour
algorithms discussed in Cook et al.
(\protect\hyperlink{ref-gt_pp_mc}{2007}). It includes methods for
geodesic interpolation and basis generation, as well as an
implementation of the simulated annealing algorithm to optimise
projection pursuit indices for the guided tour. The tour can be
displayed directly in the R graphics device, for example:

@@

\subsection{\texorpdfstring{Connecting the tour projections to D3
display using
\texttt{sendCustomMessage}}{Connecting the tour projections to D3 display using sendCustomMessage}}\label{connecting-the-tour-projections-to-d3-display-using-sendcustommessage}

@@

The new implementation of the tour uses D3 to render each projection
step returned by R, focusing on 2D projections as a test case. It does
this by drawing and re-drawing a scatterplot with dots (or circles in D3
language) and providing SVG objects for the web browser to render.

There two functions provided by the Shiny framework to transport data
be- tween R and javascript: \texttt{session\$sendCustomMessage()} in R,
and the corresponding \texttt{Shiny.addCustomMessageHandler()} in
Javascript. Whenever the former is executed in R, the latter function
will execute a code block in JS. There are many examples of such
functions being used to pass arbitrary data from an R app to a JS
front-end, few examples exist of this basic functionality to update a D3
animation in real-time.

The data format expected by D3 is in JSON format, which combines two
basic program- ming paradigms: a collection of name/value pairs, and an
ordered list of values. R's preferred data formats include data frames,
vectors and matrices. Every time a new projection has been calculated
with the tour path, the resulting matrix needs to be converted to JSON
and sent to D3. The code to send the D3 data looks like this:

@@

This code is from the observe environment from the \texttt{server.R}
file. It converts the projection matrix to JSON format, and sends it to
JavaScript with the id data. When parsed in D3 by its \texttt{data()}
method, it is converted back into a logical 2D array where the columns
are queried first, then the rows. If column names are included in the
JSON, the column indices are strings; otherwise they are integers
starting from 1. All of the code required to render the scatterplots and
legends, along with colours, is JavaScript code in the file
\texttt{d3anim.js}. In particular, the data from R is handled with the
following code:

@@

The D3 canvas makes for smooth drawing and re-drawing of the data
projections. Adding a graphical user interface (GUI) around the display
is straightforward with the shiny package, e.g.~control elements such as
stop/go, increase/decrease speed, change tour type, add/remove variables
from the mix.

The main disadvantage is that the speed is inconsistent, as server and
client play tag to keep up with each other, and the display cannot
handle many observations. Noticeable slow down was oberved with 2000
points.

@@

The \texttt{sendCustomMessage} function provides a way to share a tour
path with the D3 renderer, and embed it in a shiny GUI providing
controls such as top/go, increase/decrease speed, change tour type,
add/remove variables. However, the approach can result in inconsistent
speed which appears as jerkiness in the motion, and it isn't suitable
for large numbers of observations.

@@

\hypertarget{refs}{}
\hypertarget{ref-D3}{}
Bostock, Michael, Vadim Ogievetsky, and Jeffrey Heer. 2011. ``D3:
Data-Driven Documents.'' \emph{IEEE Transactions on Visualization and
Computer Graphics} 17 (12): 2301--9.
\url{http://vis.stanford.edu/papers/d3}.

\hypertarget{ref-shiny}{}
Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan
McPherson. 2017. \emph{Shiny: Web Application Framework for R}.
\url{https://CRAN.R-project.org/package=shiny}.

\hypertarget{ref-gt_pp}{}
Cook, Dianne, Andreas Buja, Javier Cabrera, and Catherine Hurley. 1995.
``Grand Tour and Projection Pursuit.'' \emph{Journal of Computational
and Graphical Statistics} 4 (4): 155--72.

\hypertarget{ref-gt_pp_mc}{}
Cook, Dianne, Andreas Buja, Eun-Kyung Lee, and Hadley Wickham. 2007.
``Grand Tours, Projection Pursuit Guided Tours and Manual Controls.''

\hypertarget{ref-tourrGui}{}
Huang, Bei, Dianne Cook, and Hadley Wickham. 2012. ``TourrGui: A
gWidgets Gui for the Tour to Explore High-Dimensional Data Using
Low-Dimensional Projections.'' \emph{Journal of Statistical Software} 49
(6): 1--12.

\hypertarget{ref-ihaka:1996}{}
Ihaka, Ross, and Robert Gentleman. 1996. ``R: A Language for Data
Analysis and Graphics.'' \emph{Journal of Computational and Graphical
Statistics} 5 (3): 299--314.

\hypertarget{ref-RGtk2}{}
Lawrence, Michael, and Duncan Temple Lang. 2010. ``RGtk2: A Graphical
User Interface Toolkit for R.'' \emph{Journal of Statistical Software}
37 (8): 1--52. \url{http://www.jstatsoft.org/v37/i08/}.

\hypertarget{ref-R}{}
R Core Team. 2018. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{http://www.R-project.org/}.

\hypertarget{ref-tourr}{}
Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja. 2011.
``Tourr: An R Package for Exploring Multivariate Data with
Projections.'' \emph{Journal of Statistical Software} 40 (2): 1--18.

>>>>

\abstract{%
The tourr package in R has several algorithms and displays for showing
multivariate data as a sequence of low-dimensional projections. It can
display as a movie but has no capacity for interaction, such as stop/go,
change tour type, drop/add variables. The tourrGui package provides
these sorts of controls, but the interface is programmed with the dated
RGtk2 package. This work explores using custom messages to pass data
from R to D3 for viewing, using the Shiny framework.
}

@@

The tour algorithm (Cook et al. 1995; Cook et al. 2007) is a way of
systematically generating and displaying projections of high-dimensional
spaces in order for the viewer to examine the multivariate distribution
of data. It can do this either randomly, or by picking projections
judged interesting according to some criterion or index function. The
tourr package (Wickham et al. 2011) provides the computing and display
in R (R Core Team 2018; Ihaka and Gentleman 1996) to make several types
of tours, grand, guided, little and local. The projection dimension can
be chosen between one and the number of variables in the data. The
display, though, has no capacity for interaction. The viewer can watch
the tour like a movie, but not pause it and restart, or change tour
type, or number of variables.

These interactive controls were provided with the tourrGui package
(Huang, Cook, and Wickham 2012), with was programmed with the RGtk2
package (Lawrence and {Temple Lang} 2010). This is not the toolkit of
choice today, and has been superceded with primarily web-capable tools,
like Shiny (Chang et al. 2017). To display dynamic graphics though, is
not straight-forward. This paper explains how to use D3 (Bostock,
Ogievetsky, and Heer 2011) as the display engine, in a Shiny GUI, using
custom message passing between server and client.

@@

The \pkg{tourr} package (Wickham et al. 2011) is an R implementation of
the tour algorithms discussed in Cook et al. (2007). It includes methods
for geodesic interpolation and basis generation, as well as an
implementation of the simulated annealing algorithm to optimise
projection pursuit indices for the guided tour. The tour can be
displayed directly in the R graphics device, for example:

@@

\subsection{Connecting the tour projections to D3 display using
\texttt{sendCustomMessage}}\label{connecting-the-tour-projections-to-d3-display-using-sendcustommessage}

@@

The new implementation of the tour interface uses D3 to render each
projection step returned by R, focusing on 2D projections as a test
case. It does this by drawing and re-drawing a scatterplot with dots (or
circles in D3 language) and providing SVG objects for the web browser to
render.

There are two functions provided by the Shiny framework to transport
data between R and JavaScript: \texttt{session\$sendCustomMessage()} in
R, and the corresponding \texttt{Shiny.addCustomMessageHandler()} in
JavaScript. Whenever the former is executed in R, the latter function
will execute a code block in JS. There are many examples of such
functions being used to pass arbitrary data from an R app to a JS
front-end, few examples exist of this basic functionality to update a D3
animation in real-time.

The data format expected by D3 is in JSON format, which combines two
basic programming paradigms: a collection of name/value pairs, and an
ordered list of values. R's preferred data formats include data frames,
vectors and matrices. Every time a new projection has been calculated
with the tour path, the resulting matrix needs to be converted to JSON
and sent to D3. The code to send the D3 data looks like this:

@@

This code is from the observe environment from the \texttt{server.R}
file. It converts the matrix of projected data points to JSON format,
and sends it to JavaScript with the id data. When parsed in D3 by its
\texttt{data()} method, it is converted back into a logical 2D array
where the columns are queried first, then the rows. If column names are
included in the JSON, the column indices are strings; otherwise they are
integers starting from 1. All of the code required to render the
scatterplots and legends, along with colours, is JavaScript code in the
file \texttt{d3anim.js}. In particular, the data from R is handled with
the following code:

@@

The D3 canvas makes for smooth drawing and re-drawing of the data
projections. Adding a graphical user interface (GUI) around the display
is straightforward with the Shiny package, e.g.~control elements such as
stop/go, increase/decrease speed, change tour type, add/remove variables
from the mix.

The main disadvantage is that the speed is inconsistent, as server and
client play tag to keep up with each other, and the display cannot
handle many observations. Noticeable slow down was oberved with 2000
points, the main reason being the rendering time required for the large
number of SVG circle elements. The situation can be improved when using
a single HTML5 canvas element to draw the scatter points, significantly
reducing the rendering time.

@@

The \texttt{sendCustomMessage} function provides a way to share a tour
path with the D3 renderer, and embed it in a Shiny GUI providing
controls such as stop/go, increase/decrease speed, change tour type,
add/remove variables. However, the approach can result in inconsistent
speed which appears as jerkiness in the motion, and it isn't suitable
for large numbers of observations.

@@

\hyperdef{}{refs}{}

Bostock, Michael, Vadim Ogievetsky, and Jeffrey Heer. 2011. ``D3:
Data-Driven Documents.'' \emph{IEEE Transactions on Visualization and
Computer Graphics} 17 (12): 2301--9.
\url{http://vis.stanford.edu/papers/d3}.

Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan
McPherson. 2017. \emph{Shiny: Web Application Framework for R}.
\url{https://CRAN.R-project.org/package=shiny}.

Cook, Dianne, Andreas Buja, Javier Cabrera, and Catherine Hurley. 1995.
``Grand Tour and Projection Pursuit.'' \emph{Journal of Computational
and Graphical Statistics} 4 (4): 155--72.

Cook, Dianne, Andreas Buja, Eun-Kyung Lee, and Hadley Wickham. 2007.
``Grand Tours, Projection Pursuit Guided Tours and Manual Controls.''

Huang, Bei, Dianne Cook, and Hadley Wickham. 2012. ``TourrGui: A
GWidgets GUI for the Tour to Explore High-Dimensional Data Using
Low-Dimensional Projections.'' \emph{Journal of Statistical Software} 49
(6): 1--12.

Ihaka, Ross, and Robert Gentleman. 1996. ``R: A Language for Data
Analysis and Graphics.'' \emph{Journal of Computational and Graphical
Statistics} 5 (3): 299--314.

Lawrence, Michael, and Duncan {Temple Lang}. 2010. ``RGtk2: A Graphical
User Interface Toolkit for R.'' \emph{Journal of Statistical Software}
37 (8): 1--52. \url{http://www.jstatsoft.org/v37/i08/}.

R Core Team. 2018. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{http://www.R-project.org/}.

Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja. 2011.
``Tourr: An R Package for Exploring Multivariate Data with
Projections.'' \emph{Journal of Statistical Software} 40 (2): 1--18.
